Untitled

### 2024-10-07 analysis/paper/access-ice.qmd
Adds plots for era5

```{r datasets}
datasets <- list(cdr = CDR(),
                 bt = BT(),
                 era5 = ERA5()
)
```

@@

```{r fig-hindcast-extent}
#| fig-cap: Median sea ice extent for al hindcasts initialised the first of the month for +S2 and +S1 in colours representing the start month. In black, the 
hindcast_extent |> 
  _[, median_ci(aice), by = .(forecast_time = update(forecast_time, year = 2001),
                              lag = as.numeric(as.Date(time) - forecast_time),
                              model)] |>
  _[, time := forecast_time + lag] |>
  _[, time2 := update(time, year = 2001)] |>
  _[, month := lubridate::month(forecast_time, label = TRUE)] |>
  # _[]
  ggplot(aes(time2, mid)) +
  geom_line(data = extent_climatology, 
            aes(x = as.Date(time),
                group = dataset)) +    
  geom_line(aes(color = month, group = interaction(forecast_time, time != time2))) +

@@

extent_daily |> 
  _[order(time)] |> 
  _[, dif := c(NA, diff(aice)/diff(as.numeric(as.Date(time)))), by = .(dataset)] |> 
  _[, median_ci(dif), by = .(dataset, time = update(time, year = 2000))] |> 
  ggplot(aes(time, mid)) +
  geom_hline(yintercept = 0, color = "gray50", linewidth = 0.2) +
  geom_smooth(aes(group = dataset), span = 30/366, linewidth = 0.5, color = "black", se = FALSE, n = 366)  +
  geom_smooth(data = dif_hindcast, aes(color = model), span = 30/366, linewidth = 0.5, se = FALSE, n = 366) +
  scale_x_datetime(NULL, date_breaks = "1 month", 
                   date_labels = "%b", expand = c(0, 0)) +
  scale_y_continuous(NULL, labels = labels_extent)
```

```{r climatology}
climatologies <- list(cdr = datasets[["cdr"]],
                      S2 = S2_reanalysis()) |>
  lapply(\(x) climatology(x) |>
           cdo_ymonmean() |> 
           cdo_execute()
  ) |>
  lapply(\(x) ReadNetCDF(x, "aice")) |>
  rbindlist(idcol = "model") |>
  dcast(time + xgrid + ygrid ~ model, value.var = "aice")
```

@@

climatologies |> 
  ggplot(aes(xgrid, ygrid)) +
  geom_contour_fill(aes(z = S2 - cdr, fill = after_stat(level)), 
                    breaks = AnchorBreaks(binwidth = 0.1, exclude = 0)) +
  scale_fill_divergent_discretised("ACCESS-S2 bias") +
  geomcoord_antarctica +
  geom_antarctica_fill +
  facet_wrap(~ lubridate::month(time, label = TRUE)) +
  wide_legend +
  tag_facets()
```

@@

hindcast_extent |> 
  copy() |> 
  # _[model == "S2"] |> 
  _[, aice := aice - mean(aice[year(time) %between% c(1981, 2011)]), 
    by = .(lag, model, time = update(time, year = 2000))] |> 
  _[, .(aice = mean(aice)), by = .(model, forecast_time, time = as.Date(round_date(time, "month")))] |> 
  _[, lag := months_difference(time, forecast_time) + 1] |> 
  _[lag %in% c(1, 3, 5, 7)] |> 
  ggplot(aes(time, aice)) +
  geom_line(data = monthly_extent[year(time) <= max(year(hindcast_extent$time))][dataset == "cdr"], 
            aes(group = dataset),
            color = "#1a5fb4") +
  geom_smooth(data = monthly_extent[year(time) <= max(year(hindcast_extent$time))][dataset == "cdr"], 
              aes(group = dataset), color = "#1a5fb4", se = FALSE, span = 10*12/N, n = 366) +
  geom_line()  +
  geom_smooth(se = FALSE, span = 10*12/N, n = 366, color = "black") +
  scale_y_continuous(NULL,
                     labels = labels_extent) +
  scale_x_date(NULL, expand = c(0, 0)) +
  facet_grid(lag ~ model, labeller = labeller(lag = \(x) paste0("Lag: ", x)))
```

@@

errors |> 
  _[measure == "rmse"] |>
  _[obs_dataset == "cdr"] |>
  _[, median_ci(value), 
    by = .(lag, version, measure, month(time_forecast), obs_dataset)] |> 
  ggplot(aes(lag, mid)) +
  geom_ribbon(aes(ymin = low, 
                  ymax = high, 
                  color = version, 
                  fill = version,
                  group = interaction(obs_dataset, version)),
              alpha = 0.1) +
  geom_line(aes(color = version, group = interaction(obs_dataset, version)), linewidth = 1) +
  labs(y = "RMSE", 
       x = "Lag") +
  #   scale_color_models +
  #   scale_fill_models +
  facet_wrap(~ month, labeller = labeller(month = labels_month))
```

@@

errors |> 
  _[measure == "iiee"] |>
  _[obs_dataset == "cdr"] |>
  _[, month := lubridate::month(time_forecast, label = TRUE)]  |> 
  _[, median_ci(value), 
    by = .(lag, version, measure, obs_dataset, month(time_forecast))] |> 
  ggplot(aes(lag, mid)) +
  geom_ribbon(aes(ymin = low, 
                  ymax = high, 
                  color = version, 
                  fill = version), alpha = 0.1) +
  geom_line(aes(color = version), linewidth = 1) +
  scale_y_continuous(labels = labels_extent) +
  labs(y = NULL, 
       x = "Lag") +
  #   scale_color_models +
  #   scale_fill_models +
  facet_wrap(~ month, labeller = labeller(month = labels_month))
```

@@

rbind(mean_variance = errors |>
        _[measure == "iiee"] |>
        _[obs_dataset == "cdr"] |>
        _[, var(value, na.rm = TRUE), by = .(version, lag, time_forecast)] |> 
        _[, .(value = mean(V1, na.rm = TRUE)), by = .(version, month(time_forecast), lag)],

      variance_of_mean = errors |>
        _[measure == "iiee"] |>
        _[obs_dataset == "cdr"] |>
        _[, mean(value, na.rm = TRUE), by = .(version, lag, time_forecast)] |> 
        _[, .(value = var(V1, na.rm = TRUE)), by = .(version, month(time_forecast), lag)],
      idcol = "component") |>

  ggplot(aes(lag, value)) +
  geom_line(aes(color = version), linewidth = 1) +
  scale_y_continuous("Variance", 
                     labels = labels_extent_2) +
  # scale_color_models +
  scale_x_continuous("Lag") +
  facet_grid(month ~ component, 
             labeller = labeller(month = labels_month, 
                                 component = c(mean_variance = "Mean variance",
                                               variance_of_mean = "Variance of mean"))) 
```

@@

>>>>

```{r datasets}
datasets <- list(cdr = CDR(),
                 bt = BT(),
                 era5 = ERA5()
)
which_dataset <- "cdr"
```

@@

```{r fig-hindcast-extent}
#| fig-cap: Median sea ice extent for al hindcasts initialised the first of the month for +S2 and +S1 in colours representing the start month. In black, the 
hindcast_extent |> 
  _[, median_ci(aice), by = .(forecast_time = update(forecast_time, year = 2001),
                              lag = as.numeric(as.Date(time) - forecast_time),
                              model)] |>
  _[, time := forecast_time + lag] |>
  _[, time2 := update(time, year = 2001)] |>
  _[, month := lubridate::month(forecast_time, label = TRUE)] |>
  # _[]
  ggplot(aes(time2, mid)) +
  geom_line(data = extent_climatology[dataset == which_dataset], 
            aes(x = as.Date(time),
                group = dataset)) + 
  geom_line(aes(color = month, group = interaction(forecast_time, time != time2))) +

@@

extent_daily |> 
  _[order(time)] |> 
  _[, dif := c(NA, diff(aice)/diff(as.numeric(as.Date(time)))), by = .(dataset)] |> 
  _[, median_ci(dif), by = .(dataset, time = update(time, year = 2000))] |> 
  _[dataset == which_dataset] |> 
  ggplot(aes(time, mid)) +
  geom_hline(yintercept = 0, color = "gray50", linewidth = 0.2) +
  geom_smooth(aes(group = dataset, color = dataset),
              span = 30/366, linewidth = 0.5, se = FALSE, n = 366)  +
  geom_smooth(data = dif_hindcast,
              aes(color = model), span = 30/366, linewidth = 0.5, se = FALSE, n = 366) +
  scale_x_datetime(NULL, date_breaks = "1 month", 
                   date_labels = "%b", expand = c(0, 0)) +
  scale_y_continuous(NULL, labels = labels_extent) +
  scale_color_models

```{r climatology}
climatologies <- c(datasets,
                   S2 = S2_reanalysis()) |>
  lapply(\(x) climatology(x) |>
           cdo_ymonmean() |> 
           cdo_execute()
  ) |>
  lapply(\(x) ReadNetCDF(x, "aice")) |>
  rbindlist(idcol = "model") 
```

@@

climatologies |> 
  _[model %in% c("S2", which_dataset)] |> 
  dcast(time + xgrid + ygrid ~ model, value.var = "aice") |> 
  setnames(which_dataset, "obs") |> 
  ggplot(aes(xgrid, ygrid)) +
  geom_contour_fill(aes(z = S2 - obs, fill = after_stat(level)), 
                    breaks = AnchorBreaks(binwidth = 0.1, exclude = 0)) +
  scale_fill_divergent_discretised("ACCESS-S2 bias") +
  geomcoord_antarctica +
  geom_antarctica_fill +
  facet_wrap(~ lubridate::month(time, label = TRUE)) +
  wide_legend +
  tag_facets()
```

@@

hindcast_extent |> 
  copy() |> 
  # _[model == "S2"] |> 
  _[, aice := aice - mean(aice[year(time) %between% c(1981, 2011)]), 
    by = .(lag, model, time = update(time, year = 2000))] |> 
  _[, .(aice = mean(aice)), by = .(model, forecast_time, time = as.Date(round_date(time, "month")))] |> 
  _[, lag := months_difference(time, forecast_time) + 1] |> 
  _[lag %in% c(1, 3, 5, 7)] |> 
  ggplot(aes(time, aice)) +
  geom_line(data = monthly_extent[year(time) <= max(year(hindcast_extent$time))] |> 
              _[dataset == which_dataset],
            aes(group = dataset, color = dataset)) +
  geom_smooth(data = monthly_extent[year(time) <= max(year(hindcast_extent$time))] |> 
                _[dataset == "cdr"], 
              aes(group = dataset, color = dataset),
              se = FALSE, span = 10*12/N, n = 366) +
  geom_line(aes(color = model))  +
  geom_smooth(aes(color = model), se = FALSE, span = 10*12/N, n = 366) +
  scale_y_continuous(NULL,
                     labels = labels_extent) +
  scale_x_date(NULL, expand = c(0, 0)) +
  facet_grid(lag ~ model, labeller = labeller(lag = \(x) paste0("Lag: ", x))) +
  scale_color_models
```

@@

errors |> 
  _[measure == "rmse"] |>
  _[obs_dataset == which_dataset] |>
  _[, median_ci(value), 
    by = .(lag, version, measure, month(time_forecast), obs_dataset)] |> 
  _[version == "persistence", version := obs_dataset] |> 
  ggplot(aes(lag, mid)) +
  geom_ribbon(aes(ymin = low, 
                  ymax = high, 
                  color = version, 
                  fill = version,
                  group = interaction(obs_dataset, version)),
              alpha = 0.1) +
  geom_line(aes(color = version, group = interaction(obs_dataset, version)), linewidth = 1) +
  labs(y = "RMSE", 
       x = "Lag") +
  #   scale_color_models +
  #   scale_fill_models +
  facet_wrap(~ month, labeller = labeller(month = labels_month)) +
  scale_color_models +
  scale_fill_models
```

@@

errors |> 
  _[measure == "iiee"] |>
  _[obs_dataset == which_dataset] |>
  _[, month := lubridate::month(time_forecast, label = TRUE)]  |> 
  _[, median_ci(value), 
    by = .(lag, version, measure, obs_dataset, month(time_forecast))] |> 
  ggplot(aes(lag, mid)) +
  geom_ribbon(aes(ymin = low, 
                  ymax = high, 
                  color = version, 
                  fill = version), alpha = 0.1) +
  geom_line(aes(color = version), linewidth = 1) +
  scale_y_continuous(labels = labels_extent) +
  scale_color_models +
  scale_fill_models +
  labs(y = NULL, 
       x = "Lag") +
  #   scale_color_models +
  #   scale_fill_models +
  facet_wrap(~ month, labeller = labeller(month = labels_month)) 
```

@@

rbind(mean_variance = errors |>
        _[measure == "iiee"] |>
        _[obs_dataset == which_dataset] |>
        _[, var(value, na.rm = TRUE), by = .(version, lag, time_forecast)] |> 
        _[, .(value = mean(V1, na.rm = TRUE)), by = .(version, month(time_forecast), lag)],

      variance_of_mean = errors |>
        _[measure == "iiee"] |>
        _[obs_dataset == which_dataset] |>
        _[, mean(value, na.rm = TRUE), by = .(version, lag, time_forecast)] |> 
        _[, .(value = var(V1, na.rm = TRUE)), by = .(version, month(time_forecast), lag)],
      idcol = "component") |>

  ggplot(aes(lag, value)) +
  geom_line(aes(color = version), linewidth = 1) +
  scale_y_continuous("Variance", 
                     labels = labels_extent_2) +
  # scale_color_models +
  scale_x_continuous("Lag") +
  scale_color_models +
  facet_grid(month ~ component, 
             labeller = labeller(month = labels_month, 
                                 component = c(mean_variance = "Mean variance",
                                               variance_of_mean = "Variance of mean"))) 
```

@@

```{r dataset_era}
which_dataset <- "era5"
```

```{r}
<<fig-hindcast-extent>>
```

```{r}
<<fig-mean-growth>>
```

```{r}
<<fig-bias>>
```

```{r}
<<<fig-extent-anom>>
```

```{r}
<<fig-rmse>>
```

```{r}
<<fig-iiee>>
```

```{r}
<<fig-iiee-variance>>
```
